---
title: "Bios"
author: "Ranxuan Li, Yongxian Zhang, Sinuo Lu"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r global_setup, include=FALSE}

## we can set the global default behavior of the code chunks using the following command. You can find out more about additionalguidance on the use of the opts_chunk$set() function of the knitr package at https://yihui.org/knitr/options/#chunk-options.

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, comment="")

## You can load the R packages that you'll use here as well. 
library(here)
library(tidyverse)
library(dplyr)
library(car)
library(stringr)
library(ggplot2)
library(gt)
library(gtsummary)
library(flextable)
```

# Data Pre-processing and Testing

## Numerical variables

```{r data_prep, echo=FALSE, eval=FALSE}
data_raw = read.csv("kidney_disease.csv")

data_deblank = data_raw %>%
  mutate(across(everything(), ~na_if(trimws(.), ""))) %>%
  select(age, sg, al, bgr, bu, sc, hemo, pcv, htn, dm, ane, classification) 
data_left = sapply(data_deblank[,1:8], as.numeric) 
data_right = data_deblank[, -1:-8]
data = cbind(data_left, data_right)

#exclude hidden spaces/tabs
data$classification = str_trim(data$classification)

#fill num vari NA with median and chara NA with mode #is.numeric(data[[1]])
get_mode <- function(x) {
  ux <- na.omit(unique(x))
  ux[which.max(tabulate(match(x, ux)))]
}
data <- data %>%
  mutate(across(where(is.numeric), ~replace_na(., median(., na.rm=TRUE)))) %>%
  mutate(across(where(is.character), ~replace_na(., get_mode(.))))

```

### Test whether equal variance

```{r explore, echo=FALSE, eval=FALSE}

# We use levene test to see whether inter-group variance is different to determine which testing method for use.

leveneTest(sg ~ classification, data = data) #, center = "median")
leveneTest(al ~ classification, data = data)
leveneTest(bgr ~ classification, data = data)
leveneTest(bu ~ classification, data = data)
leveneTest(sc ~ classification, data = data)
leveneTest(hemo ~ classification, data = data)
leveneTest(pcv ~ classification, data = data)
```

### test whether normal distribution

```{r}
# we plot the data to see normality
ggplot(data=data, aes(sample = al)) +
  stat_qq() + 
  stat_qq_line() +
  facet_wrap(~ classification)

ggplot(data=data, aes(x = bgr)) +
  geom_histogram()+
  facet_wrap(~ classification)


```


### data not normal, t-test assumption violated
```{r}
# Data is not normally distributed, so t-tests are not accurate

# if we use only complete observations
#data = data[complete.cases(data), ]

# t.test(not equal var)
t.test(sg ~ classification, data=data, na.exclude=TRUE, var.equal = F)
t.test(al ~ classification, data=data, na.exclude=TRUE, var.equal = F)
t.test(bgr ~ classification, data=data, na.exclude=TRUE, var.equal = F)
t.test(bu ~ classification, data=data, na.exclude=TRUE, var.equal = F)
t.test(sc ~ classification, data=data, na.exclude=TRUE, var.equal = F)
t.test(hemo ~ classification, data=data, na.exclude=TRUE, var.equal = F)
t.test(pcv ~ classification, data=data, na.exclude=TRUE, var.equal = F)
```




### not normal, use wilcox.test()
```{r}
#we use wilcox.test in this situation
wilcox.test(sg ~ classification, data=data)
wilcox.test(al ~ classification, data=data)
wilcox.test(bgr ~ classification, data=data)
wilcox.test(bu ~ classification, data=data)
wilcox.test(sc ~ classification, data=data)
wilcox.test(hemo ~ classification, data=data)
wilcox.test(pcv ~ classification, data=data)
wilcox.test(age ~ classification, data=data)

```


## Categorical variables.

```{r}
# for categorical values, we use chiaq tests
htn_table = table(data$htn, data$classification)
dm_table = table(data$dm, data$classification)
ane_table = table(data$ane, data$classification)

h = chisq.test(htn_table, correct=FALSE)
d = chisq.test(dm_table, correct=FALSE)
a = chisq.test(ane_table, correct=FALSE)

h;d;a

h$expected
d$expected
a$expected
# all expectations > 5, chisq test valid
```








## Categorical variables.

```{r}
htn_table = table(data$htn, data$classification)
dm_table = table(data$dm, data$classification)
ane_table = table(data$ane, data$classification)

h = chisq.test(htn_table, correct=FALSE)
d = chisq.test(dm_table, correct=FALSE)
a = chisq.test(ane_table, correct=FALSE)

h;d;a

h$expected
d$expected
a$expected
# all expectations > 5
```

# EDA
### Summary data analysis
```{r summ}
cat_vars <- c("rbc", "pc", "pcc", "ba", 
              "htn", "dm", "cad", "appet", "pe", "ane", "classification")

data_sum <- data_deblank_complete %>% 
  select(-"id") %>%
  mutate(across(all_of(cat_vars), as.factor))
```

```{r groupedsum}
data_sum2 <- data_sum %>%
  mutate(across(where(is.numeric), ~replace_na(., median(., na.rm=TRUE)))) %>%
  mutate(across(where(is.character), ~replace_na(., mode(.))))

# produce summary statistics for the variables
summary_table <- data_sum %>%
  tbl_summary(
    by = classification,
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    missing = "no"
  )%>%
  add_p() %>% 
  add_overall() %>%
  bold_p() %>%                 
  modify_header(label = "**Variable**")

summary_table

# to output the table in a docx
summary_table %>% 
  as_flex_table() %>% 
  flextable::save_as_docx(path = "CKD_table1.docx")

```



The baseline characteristics show that CKD patients differ substantially from non-CKD participants in almost every clinical, laboratory, and demographic variable. Continuous kidney function markers (serum creatinine, blood urea, hemoglobin, PCV, sodium) display large and statistically significant differences. CKD patients presented with markedly higher rates of hypertension, diabetes, proteinuria, glycosuria, urinary abnormalities, anemia, and poor appetite.

Overall, the table clearly demonstrates that CKD patients exhibit the classical clinical and biochemical signatures of renal impairment.



### Correlation plot

```{r corrplot}
data_deblank_complete = data_raw %>%
  mutate(across(everything(), ~na_if(trimws(.), ""))) 
data_num = sapply(data_deblank_complete[,c(2:6,11:19)], as.numeric) 
data_chara = data_deblank_complete[, c(1,7:10,20:26)]
data_deblank_complete = cbind(data_num, data_chara)

# choose only numeric variables for corrleation plotting

data_num <- data_num %>%
  as.data.frame()%>%
  mutate(across(where(is.numeric), ~replace_na(., median(., na.rm=TRUE))))

data_corr = data_num 
colnames(data_corr) = c("Age", "Blood presure", "Specific Gravity (urine)", "Albumin in urine", "Sugar in urine",
                    "Blood Glucose Random", "Blood Urea", "Serum Creatinine","Sodium in blood","Potassium in blood",
                    "Hemoglobin level","Packed Cell Volume (Hematocrit)","White Blood Cell Count","Red Blood Cell Count")
# Correlation on complete cases of X
cors = cor(data_corr)
#round(cors, 2)

# Visualize (corrplot)
corrplot::corrplot(cors, method = "color", type = "upper",
                   number.cex = .6, tl.cex = 0.7,
                   tl.col = "black", tl.srt = 45, diag = FALSE)

```
The correlation heatmap reveals clinically consistent clusters among the biochemical and clinical markers. Serum creatinine and blood urea show a strong positive correlation, reflecting their shared role as indicators of declining glomerular filtration rate. Hemoglobin and packed cell volume also correlate strongly, forming a distinct anemia-related cluster, consistent with the loss of erythropoietin production in chronic kidney disease.

Blood glucose and urine sugar demonstrate moderate positive correlation, indicating a metabolic cluster related to diabetes mellitus, a major risk factor for CKD. In contrast, electrolytes (sodium and potassium) and white blood cell count exhibit weak correlations with most variables, suggesting limited discriminative value for CKD in this dataset.

Overall, the correlation plot highlights three clinically relevant domains—renal filtration impairment, anemia, and diabetes, each strongly associated with CKD and consistent with the variables selected for further analysis.

### Heat map

```{r}
data_binary <- data_chara %>%
  select(-1)%>%
  mutate(across(everything(), ~ as.numeric(as.factor(.x)) - 1))

data_heatmap = cbind(data_num, data_binary)
cor_mat <- cor(data_heatmap, method = "pearson", use = "pairwise.complete.obs")
cor_long <- cor_mat %>%
  as.data.frame() %>%
  mutate(var1 = row.names(cor_mat)) %>%
  pivot_longer(
    cols = -var1,
    names_to = "var2",
    values_to = "correlation"
  )

ggplot(cor_long, aes(x = var1, y = var2, fill = correlation)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient2(
    low = "cyan",
    mid = "blue",
    high = "red",
    midpoint = 0,
    name = "Pearson Correlation"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_blank()
  )

```


### Age as a potential confounder

Both the summary table and the correlation matrix indicate that age is strongly associated with CKD status (p < 0.001), with CKD patients being substantially older on average. The corrplot further shows that age is positively correlated with blood pressure and negatively correlated with hemoglobin levels. These variables are also significantly associated with CKD.

Because age simultaneously influences the exposure variables (e.g., hemoglobin, hypertension) and the outcome (CKD), and is unequally distributed between groups, age satisfies the criteria for being a confounder. Therefore, the observed association between hemoglobin or hypertension and CKD may be partially explained by differences in age distribution.

To further investigate whether age acts as a confounder in the association between hemoglobin, hypertension, and CKD status, we examined the relationships between age and these variables. 

```{r boxplothpn}
# see whether age differs between CKD groups and non-ckd groups
data %>%
  ggplot(aes(classification,age, colour = classification)) +
  geom_boxplot() +
  labs(title = "Boxplots of Age by CKD Status",
       color = "CKD")+
  theme_minimal()+
  theme(text = element_text(face = "bold"),
        plot.title = element_text(hjust = 0.5))

```

```{r lm}
# see if age is correlated with Hemoglobin Level
ggplot(data, aes(x = age, y = hemo)) +
 geom_point(alpha = 0.5) +
 geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, color = "red") +
 labs(x = "Age", y = "Hemoglobin Level",
 title = "Relationship Between Patient Age and Hemoglobin Level with Linear Regression") +
 theme_minimal()+
 theme(text = element_text(face = "bold"),
 plot.title = element_text(hjust = 0.5))
```

The scatterplot of hemoglobin versus age demonstrates a mild negative linear trend, indicating that hemoglobin levels tend to decrease slightly with increasing age. Although the variability is large, this pattern is consistent with age-related declines in erythropoietin production and general physiological changes in older adults.

In contrast, the boxplot comparing age distributions by hypertension status shows a pronounced difference: hypertensive patients are substantially older than non-hypertensive patients. This strong age gradient reflects the well-established epidemiological reality that hypertension is primarily an age-associated condition.

Both of the figures suggest that age is a confounding variable in our analysis. Age influences both hemoglobin levels and hypertension prevalence and may therefore act as a confounder in the relationship between these variables and chronic kidney disease.

```{r wctest}
wilcox.test(age ~ htn, data=data)
wilcox.test(age ~ classification, data=data)

summary(lm(age~sc, data=data))

```







# linear regression Analysis

```{r}
library(caret)#
```

## Data preparation

```{r}
# data preparation for the second worker

df = read.csv("kidney_clean.csv")
#glimpse(df)
#str(df)
head(df)
```

```{r}
sum(duplicated(df))
# no duplicated values

#df[duplicated(df),]
```

```{r ,warning=FALSE,message=FALSE}

#df <- subset(df, select = -id)#please run this for one first time
#df <- df[, !names(df) %in% c("id")]
#str(df)
#colnames(df)




num_cols <- c(
  "Age.of.the.patient",
  "Blood.Pressure",
  "Specific.Gravity..urine.",
#  "Albumin.in.urine",
#  "Sugar.in.urine",
  "Blood.Glucose.Random",
  "Blood.Urea",
  "Serum.Creatinine",
  "Sodium.in.blood",
  "Potassium.in.blood",
  "Hemoglobin.level",
  "Packed.Cell.Volume..Hematocrit.",
  "White.Blood.Cell.Count",
  "Red.Blood.Cell.Count"
)

df[num_cols] <- lapply(df[num_cols], function(x) as.numeric(x))


str(df)
cat_cols <- names(df)[sapply(df, is.character) | sapply(df, is.factor)]
#num_col <- names(df)[!(sapply(df, is.character) | sapply(df, is.factor))]
#
cat_cols
#num_col

#summary(df)
```

```{r}
df <- df[-1,]

df$`Diabetes.Mellitus` <- gsub("\\\tno", "no", df$`Diabetes.Mellitus`)
df$`Diabetes.Mellitus` <- gsub("\\\tyes", "yes", df$`Diabetes.Mellitus`)

df$`Coronary.Artery.Disease` <- gsub("\\\tno", "no", df$`Coronary.Artery.Disease`)
df$`Coronary.Artery.Disease` <- gsub("\\\tyes", "yes", df$`Coronary.Artery.Disease`)

df$Diagnosis.label  <- gsub("ckd\\\t", "ckd", df$Diagnosis.label )
#summary(df)
```

```{r}

colSums(is.na(df))
```

```{r}
# Replace NA values in a numeric vector with randomly sampled existing (non-NA) values from the same variable.

#random -- NA--numeric
Random_value_Imputation <- function(x, seed=42) {
  set.seed(seed)
  na_index <- is.na(x)
  x[na_index] <- sample(x[!na_index], sum(na_index), replace=TRUE)
  return(x)
}# "Albumin.in.urine"                "Sugar.in.urine" 

#replaces all NA values in a categorical vector with the mode
#mode -- NA--cate
impute_mode <- function(x, seed=42) {
  set.seed(seed)
  mode_val <- names(sort(table(x), decreasing = TRUE))[1]
  x[is.na(x)] <- mode_val
  return(x)
}

#num_cols Loop over all numeric columns and perform random-value imputation
#cat_cols Loop over all categorical columns and impute missing values using the mode
for (col in num_cols) {
  df[[col]] <- Random_value_Imputation(df[[col]], seed=203)
}
for (col in cat_cols) {
  df[[col]] <- impute_mode(df[[col]], seed=203)
}
```

## corelation heatmap

```{r message=FALSE, warning=FALSE}
library(corrplot)
library(dplyr)
library(ggplot2)
library(reshape2)

num_data <- df[, num_cols]

# ensure all variables are numeric
num_data <- mutate_all(num_data, function(x) as.numeric(as.character(x)))

# correlation matrix calculation, use complete.obs only
corr_matrix <- cor(num_data, use = "complete.obs")
# long-format for plotting
corr_melt <- melt(corr_matrix)

# plotting
ggplot(corr_melt, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "#2E86AB", mid = "white", high = "#E74C3C",
                       midpoint = 0, limit = c(-1, 1), name = "Correlation") +
  theme_minimal(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1),
        panel.grid = element_blank(),
        axis.title = element_blank()) +
  coord_fixed() +
  geom_text(aes(label = sprintf("%.2f", value)), size = 1.5, color = "black") +
  labs(title = "Correlation Heatmap of Numeric Variables")
```

```{r}
summary(df)

#colSums(is.na(df))
head(df)
#df<-subset(df,select = -Patient.ID)
cat_cols <- setdiff(cat_cols, "Patient.ID")


# Perform label encoding for categorical variables
df$Hypertension <- as.integer(factor(df$Hypertension))
df$Diabetes.Mellitus <- as.integer(factor(df$Diabetes.Mellitus))
df$Coronary.Artery.Disease <- as.integer(factor(df$Coronary.Artery.Disease))
df$Appetite <- as.integer(factor(df$Appetite))
df$Pedal.Edema..swelling.of.legs. <- as.integer(factor(df$Pedal.Edema..swelling.of.legs.))
df$Anemia <- as.integer(factor(df$Anemia))
df$Red.Blood.Cells.in.urine <- as.integer(factor(df$Red.Blood.Cells.in.urine))
df$Albumin.in.urine <- as.integer(factor(df$Albumin.in.urine))
df$Sugar.in.urine <- as.integer(factor(df$Sugar.in.urine))
df$Pus.Cells.in.urine <- as.integer(factor(df$Pus.Cells.in.urine))
df$Pus.Cell.Clumps <- as.integer(factor(df$Pus.Cell.Clumps))
df$Bacteria.in.urine <- as.integer(factor(df$Bacteria.in.urine))

#df$packed_cell_volume <- as.integer(factor(df$packed_cell_volume))
#df$white_blood_cell_count <- as.integer(factor(df$white_blood_cell_count))

#df$red_blood_cell_count <- as.integer(factor(df$red_blood_cell_count))
df$ Diagnosis.label <- as.integer(factor(df$ Diagnosis.label))
cat_cols

str(df)
```

```{r}
sapply(df, is.numeric)
```

```{r warning=FALSE,message=FALSE}

# Train–test split (80% training, 20% testing)
library(caret)
# Create training and testing subsets
index <- createDataPartition(df$ Diagnosis.label, p = 0.8, list = FALSE)
train_data <- df[index, ]
test_data <- df[-index, ]

# Convert Diagnosis.label to binary (0/1)
train_data$ Diagnosis.label <- ifelse(train_data$ Diagnosis.label == 2, 1, 0)
test_data$ Diagnosis.label <- ifelse(test_data$ Diagnosis.label == 2, 1, 0)

# Create predictor matrices (X) and outcome vectors (y)
X_train <- as.matrix(train_data[, -which(names(train_data) == "Diagnosis.label")])
X_test  <- as.matrix(test_data[, -which(names(test_data) == "Diagnosis.label")])
y_train <- train_data$ Diagnosis.label
y_test  <- test_data$ Diagnosis.label

summary(train_data)

```


```{r}
target <- "Serum.Creatinine"
# Remove the target from numeric columns so we only plot predictors
num_cols_plot <- setdiff(num_cols, target)

plot_list <- lapply(num_cols_plot, function(v){
  ggplot(df, aes_string(x = v, y = target)) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    labs(title = paste("Serum Creatinine vs", v),
         x = v,
         y = "Serum Creatinine")
})

plot_list





```

## 2.Relationship between kidney function and hemoglobin

Q1:Is kidney function associated with hemoglobin level in CKD patients? --Kidney dysfunction is significantly associated with lower hemoglobin.

```{r}
write.csv(df, "kidney_clean.csv", row.names = FALSE)

# Plot: Serum Creatinine vs Hemoglobin Level
ggplot(df, aes(x = Serum.Creatinine, y = Hemoglobin.level)) +
geom_point(alpha = 0.6) +
geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal()+
labs(title = "Relationship between Creatinine and Hemoglobin.level",
x = "Serum Creatinine", y = "Hemoglobin.level")



```

Extrema of Serum Creatinine strongly leverage and influence linear regression, causing the blue regression line to be "skewed," a very typical problem in CKD data.

Instead of crudely deleting data, we should handle outliers using statistically sound methods, including sensitivity analysis.

####Step 1: Quantify outliers & leverage using formal metrics 1.1 Calculate Cook's Distance (impact point)

```{r}

fit_sc <- lm(Hemoglobin.level ~ Serum.Creatinine, data = df)

# Cook's distance
cook <- cooks.distance(fit_sc)

# potential influencial points
which(cook > 4/length(cook))

plot(fit_sc, which = 5)

```

1.2 check point leverage

```{r}
# Calculate leverage (hat values)
lev <- hatvalues(fit_sc)

# Plot leverage values
plot(lev, type = "h",
     main = "Leverage Values",
     ylab = "Leverage")
# Add horizontal threshold line; Rule of thumb: 2 * average leverage
abline(h = 2 * mean(lev), col = "red", lty = 2)

```

#### check the distribution of Derum Creatinine

```{r}
hist(df$Serum.Creatinine,breaks = 30)


```

### Medically appropriate management (avoiding arbitrary deletion of data)

In CKD, Creatinine levels are inherently severely right-biased, so recommended approaches include: ✅ Method A: Taking the logarithm of Creatinine

```{r}
df <- df %>%
  mutate(log_creatinine = log(Serum.Creatinine + 0.1))

ggplot(df, aes(x = log_creatinine, y = Hemoglobin.level)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  theme_minimal()+
  labs(title = "Hemoglobin vs log(Serum Creatinine)",
       x = "log(Serum Creatinine)",
       y = "Hemoglobin level")
```

Scatterplots revealed several extreme serum creatinine values exerting high leverage on the linear regression line. Given the known right-skewed distribution of creatinine in CKD patients, a log-transformation was applied to reduce the influence of extreme observations and improve model stability. After transformation, the relationship between creatinine and hemoglobin appeared more linear and less driven by a small number of influential cases, resulting in a more reliable estimation of the association.

```{r}
fit <- lm(Hemoglobin.level ~ log_creatinine, data = df)

summary(fit)


fit$coefficients   # intercept & slope


```

residuals sum to 0

mean(fitted) = mean(observed)

```{r}


sum(fit$residuals)         # should be ~ 0
mean(fit$fitted.values)
mean(df$Serum.Creatinine)


```

## Regression DIagnosis

```{r}

plot(fit$fitted.values, fit$residuals,
xlab="Fitted values", ylab="Residuals",
main="Residuals vs Fitted", pch=20, col="steelblue")
abline(h=0,lwd=2,lty=2)

par(mfrow = c(2, 2))      # 2x2 排版
plot(fit) 

```

```{r}

qqnorm(fit$residuals, pch=20, col="darkred")
qqline(fit$residuals, col="blue", lwd=2)



```

```{r}
plot(fit, which = 1)
plot(fit, which = 2)
plot(fit, which = 4)
plot(fit, which = 3)


plot(fit, which = 5)


```

## R2 explaination

```{r}

summary(fit)$r.squared


summary(df)
```

### 3.Crude association between age and hemoglobin

Q2:Is age associated with hemoglobin after accounting for kidney disease severity? --Age appears to be negatively associated with hemoglobin in the crude model.

#### Crude model with only age

```{r}
fit_age <- lm(Hemoglobin.level ~ Age.of.the.patient, data = df)
summary(fit_age)
# Scatterplot with fitted regression line
ggplot(df, aes(x = Age.of.the.patient, y = Hemoglobin.level)) +
  geom_point(alpha = 0.6, color = "gray30") +
  geom_smooth(method = "lm", se = FALSE, color = "darkred", linewidth = 1.2) +
  theme_minimal() +
  labs(
    title = "Crude Association between Age and Hemoglobin Level",
    subtitle = "Unadjusted linear regression model",
    x = "Age of Patient (years)",
    y = "Hemoglobin Level (g/dL)"
  )

```

### (2)Age effect after adjusting for renal function

```{r}

df$Age_std   <- as.numeric(scale(df$Age.of.the.patient))
df$logC_std  <- as.numeric(scale(df$log_creatinine))

# Fit linear model with standardized predictors
fit_std <- lm(Hemoglobin.level ~ Age_std + logC_std, data = df)
summary(fit_std)
library(ggeffects)

# Obtain adjusted effect predictions using ggpredict
p_age <- ggpredict(fit_std, terms = "Age_std")
p_age$variable <- "Age"

p_cre <- ggpredict(fit_std, terms = "logC_std")
p_cre$variable <- "Creatinine"

pp <- rbind(p_age, p_cre)
library(ggplot2)

ggplot() +
  # original standardized scatter points  
  geom_point(data = df, 
             aes(x = Age_std, y = Hemoglobin.level),
             alpha = 0.35, color = "gray40") +
  geom_point(data = df, 
             aes(x = logC_std, y = Hemoglobin.level),
             alpha = 0.35, color = "gray70") +

  # adjusted regression lines
  geom_line(data = pp,
            aes(x = x, y = predicted, color = variable),
            size = 1.25) +
  geom_ribbon(data = pp,
              aes(x = x, ymin = conf.low, ymax = conf.high, fill = variable),
              alpha = 0.15, color = NA) +

  theme_minimal() +
  labs(
    title = "Adjusted Effects of Age and Creatinine on Hemoglobin",
    subtitle = "Partial regression lines after standardizing predictors",
    x = "Standardized Predictor Value",
    y = "Hemoglobin Level (g/dL)",
    color = "Effect",
    fill = "Effect"
  )



```

To account for the effect of kidney function, a multivariable linear regression model was fitted with hemoglobin level as the outcome and both serum creatinine and age as predictors. After adjustment for creatinine, the association between age and hemoglobin was substantially attenuated compared to the crude model, indicating that the apparent relationship between age and hemoglobin is partly mediated or confounded by renal function severity. This suggests that kidney dysfunction plays a more central role in explaining anemia in CKD patients than age alone.

```{r}
# Generate adjusted predictions for Age 
pred <- ggpredict(fit_adj, terms = "Age.of.the.patient")

# Plot adjusted partial effect of Age on Hemoglobin
ggplot(pred, aes(x = x, y = predicted)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
  theme_minimal() +
  labs(
    title = "Adjusted Effect of Age on Hemoglobin",
    subtitle = "Controlling for Serum Creatinine",
    x = "Age (years)",
    y = "Predicted Hemoglobin (g/dL)"
  )
```

The line is slightly upward. From young to old, the predicted change in Hb is very small (approximately 0.3 g/dL). After controlling for kidney function levels, the effect of age on hemoglobin is very weak, almost clinically insignificant.

### Simpson Paradox

#### Effect modification by hypertension

Q3:Does hypertension modify the relationship between age and hemoglobin? --· However, this association is distorted after stratification by hypertension. · · This demonstrates Simpson’s paradox caused by hypertension as a confounder, reflecting its dual association with both advanced age and anemia severity.

```{r}
df_plot <- df %>%
  filter(Hypertension %in% c(2, 3)) %>%
  mutate(Hypertension = factor(Hypertension,
                               levels = c(2, 3),
                               labels = c("No Hypertension", "Hypertension")))

## overall fitting curve

p_overall <- ggplot(df_plot, aes(x = Age.of.the.patient,
                                 y = Hemoglobin.level)) +
  geom_point(aes(color = Hypertension), alpha = 0.6, size = 2) +
  # 整体一条回归线（不分组）
  geom_smooth(method = "lm", se = FALSE,
              color = "black", linewidth = 1.2) +
  scale_color_brewer(palette = "Set1",
                     name = "Hypertension Status") +
  labs(
    title = "Age vs Hemoglobin (Overall Fit)",
    x = "Age of Patient (years)",
    y = "Hemoglobin Level (g/dL)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

## fitting by groups

p_grouped <- ggplot(df_plot, aes(x = Age.of.the.patient,
                                 y = Hemoglobin.level,
                                 color = Hypertension)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  scale_color_brewer(palette = "Set1",
                     name = "Hypertension Status") +
  labs(
    title = "Age vs Hemoglobin Stratified by Hypertension",
    x = "Age of Patient (years)",
    y = "Hemoglobin Level (g/dL)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p_overall
p_grouped

  

```

## simpson paradox calculation

## data preparation

```{r}

library(dplyr)

df_plot <- df %>%
  filter(Hypertension %in% c(2, 3)) %>%
  mutate(
    Hypertension = factor(
      Hypertension,
      levels = c(2, 3),
      labels = c("No Hypertension", "Hypertension")
    )
  )

table(df_plot$Hypertension)
summary(df_plot$Age.of.the.patient)
summary(df_plot$Hemoglobin.level)




```

## fit the model

```{r}

# Overall linear regression: Hemoglobin ~ Age
fit_all <- lm(Hemoglobin.level ~ Age.of.the.patient, data = df_plot)
summary(fit_all)
overall_slope  <- coef(fit_all)["Age.of.the.patient"]
overall_pvalue <- summary(fit_all)$coefficients["Age.of.the.patient", "Pr(>|t|)"]

overall_slope
overall_pvalue

```

```{r}


# Hypertension = "Hypertension" 
fit_htn <- lm(Hemoglobin.level ~ Age.of.the.patient,
              data = df_plot %>% filter(Hypertension == "Hypertension"))
summary(fit_htn)

htn_slope  <- coef(fit_htn)["Age.of.the.patient"]
htn_pvalue <- summary(fit_htn)$coefficients["Age.of.the.patient", "Pr(>|t|)"]

htn_slope
htn_pvalue

# Hypertension = "No Hypertension" 
fit_nohtn <- lm(Hemoglobin.level ~ Age.of.the.patient,
                data = df_plot %>% filter(Hypertension == "No Hypertension"))
summary(fit_nohtn)

nohtn_slope  <- coef(fit_nohtn)["Age.of.the.patient"]
nohtn_pvalue <- summary(fit_nohtn)$coefficients["Age.of.the.patient", "Pr(>|t|)"]

nohtn_slope
nohtn_pvalue

```

However, this association is distorted after stratification by hypertension. This demonstrates Simpson’s paradox caused by hypertension as a confounder, reflecting its dual association with both advanced age and anemia severity.

#### Clinical interpretation

These findings highlight that hypertension acts as a confounding variable in the relationship between age and hemoglobin. Ignoring hypertension status would lead to the misleading conclusion that increasing age directly worsens anemia severity, whereas in reality, the apparent decline is driven by the disproportionate representation of hypertensive individuals in older age groups. Therefore, hypertension must be explicitly accounted for when evaluating age-related changes in hemoglobin among patients with chronic kidney disease.

### Validation of Age as the Confounder of Hypertension

New Module: Validating Age as the Confounder of Hypertension Step 1: Correlation between Age and Hypertension (Numerical + Visualization) 1.1 Spearman/Pearson Correlation

```{r}
# binary conversion for interpreting
df$Hypertension_bin <- ifelse(df$Hypertension == 3, 1, 0)

cor.test(df$Age.of.the.patient,
         df$Hypertension_bin,
         method = "spearman")

t.test( Age.of.the.patient~ Hypertension_bin, data = df)


```

```{r}

ggplot(df_plot, aes(x = factor(Hypertension),
               y = Age.of.the.patient,
               fill = factor(Hypertension))) +
  geom_boxplot(alpha = 0.7) +
  scale_x_discrete(labels = c("No HTN", "HTN")) +
  labs(
    title = "Distribution of Age by Hypertension Status",
    x = "Hypertension Status",
    y = "Age (years)"
  ) +
  theme_minimal()


```

Boxplot shows that the age distribution in the hypertension group is significantly higher than in the non-hypertension group, with both the median and quartiles shifting to the right, further indicating that hypertension is more common in the elderly.

### Step 2: Differences in Hemoglobin between Hypertension Groups

```{r}

ggplot(df_plot, aes(x = factor(Hypertension),
               y = Hemoglobin.level,
               fill = factor(Hypertension))) +
  geom_boxplot(alpha = 0.7) +
  scale_x_discrete(labels = c("No HTN", "HTN")) +
  labs(
    title = "Hemoglobin Level by Hypertension Status",
    x = "Hypertension Status",
    y = "Hemoglobin (g/dL)"
  ) +
  theme_minimal()

t.test(Hemoglobin.level ~ Hypertension_bin, data = df)

```

Compared to those without hypertension, patients in the hypertension group had significantly lower hemoglobin levels (p \< 0.05), suggesting that hypertension is associated with the degree of anemia.

### Confounding Logic Summary

Age satisfies the criteria for a confounding variable in the relationship between hypertension and hemoglobin:

Age is significantly associated with Hypertension, with hypertensive patients being notably older on average.

Age is also negatively associated with Hemoglobin level in the crude analysis.

After adjusting for renal function (serum creatinine), the apparent relationship between Age and Hemoglobin becomes substantially attenuated, indicating that the crude association was partly driven by the uneven age distribution across hypertension groups.

Therefore, Age acts as a confounder by influencing both the exposure (Hypertension) and the outcome (Hemoglobin), and failing to adjust for Age would result in a distorted interpretation of the hypertension–anemia relationship.
